---
title: "STA240 Spring 2025, Lab 1"
subtitle: Due Friday January 17 at 5:00 PM
format: html
editor: visual
---

## Important techno-babble

Throughout this course, it may serve you well to familiarize yourself with the following:

-   **R/RStudio** for coding;
-   **LaTeX** (or $\LaTeX$, if you're cool) for typesetting math;
-   **Quarto** for synthesizing your expertly written code and your beautifully typeset math in a wonderfully legible PDF format!

Of these three tools, we will focus the most on R/RStudio during lab time, though LaTeX and Quarto will certainly make incidental appearances.

To use R/RStudio, you can either use the [Duke Container Manager](https://cmgr.oit.duke.edu/containers) or download the necessary applications [here](https://posit.co/downloads/). Note: The R/RStudio in the containers comes with `TinyTeX` already downloaded. So, if you download the application and want to use LaTeX in your files, run the following command (and whatever subsequent commands you are prompted to run):

```{r}
#| eval: false
tinytex::install_tinytex()
```

For a quick guide to useful LaTeX symbols, check out [this OEIS wiki](https://oeis.org/wiki/List_of_LaTeX_mathematical_symbols) or [this Rice page](https://www.cmor-faculty.rice.edu/~heinken/latex/symbols.pdf).

To learn more about Quarto, visit <https://quarto.org>.

## Skillz

Besides helping you exercise your probability and statistics muscles, each lab will help you develop your understanding of and comfort with various coding concepts and skills. Although the implementation will vary from language to language, the underlying structure will remain the same!

This lab exercises the following coding concepts:

-   Vectors/Arrays of numbers and strings;
-   `for`-loops;
-   Boolean logic and operations;
-   Base `R` functions like `mean()` and `sample()`.

## How does one count? (I do it.)

### Coin Toss

We first consider the flip of a fair coin. That is, when flipped, the coin will equiprobably come up heads or tails. Suppose we try to illustrate the fairness of a coin by physically flipping it a bunch of times. Very quickly, this becomes tedious and labor-intensive; surely, no reasonable/sane person would want to do this by hand...

(Though, on the off chance that [somebody actually did this](https://arxiv.org/pdf/2310.04153), at least you might get co-authorship and recognition from [various](https://www.scientificamerican.com/article/scientists-destroy-illusion-that-coin-toss-flips-are-50-50/) sectors of [media](https://youtu.be/54pmrsCGt4o?t=239).)

At any rate, R provides a much more enticing, efficient alternative to being one of about $50$ people flipping coins thousands of times.

First, let's define a vector with the sample space -- all possible outcomes -- of a single coin flip. We notate heads as `H` and tails as `T`, and we call the vector `sides_of_coin`.

```{r coinsampspace}
sides_of_coin <- c("H", "T")
```

To simulate an actual coin flip, R has this nifty built-in function called `sample()`.

#### The `sample()` function

The `sample()` function can take many arguments: a vector of elements (the sample space), a positive integer (the size), a Boolean (with or without replacement), and a vector of probabilities. So, a call to `sample()` will return a random sample of the specified size from the sample space, either with or without replacement, according to the vector of probabilities.

NOTE: The `replace` Boolean defaults to `FALSE` (without replacement), and the `prob` vector defaults to equal weighting among the elements in the sample space.

#### Sample flips

We can perform a single coin flip with the following function call.

```{r sampletoss}
sample(sides_of_coin, 1)
```

If we wanted to simulate multiple coin flips in the same vector, `sample()` lets us do that as well! Below is an example of a simulation of $10$ coin flips.

```{r sample10toss}
sample(sides_of_coin, 10, replace = TRUE)
```

We can also use a different tool to simulate multiple fair coin flips: a `for`-loop.

#### `for`-loops

A `for`-loop is a great way to perform the same set of commands a fixed, known number of times or to perform the same set of commands on the different elements in a vector. The bare bones of the `for`-loop are given below.

```{r emptyloop}
#| eval: FALSE
for(){
  
}
```

In the parentheses `()`, you define the variable and values of iteration; in the brackets `{}`, you list the commands to be performed in each iteration.

#### Simulating the flips

We will simulation $5,000$ independent flips of a fair coin and track whether each flip came up heads. We do this by initializing a $5,000$-element vector of zeroes and replacing each zero with a `TRUE` or `FALSE` according to the result of the current coin flip. You may find the corresponding code below.

```{r coinloop}
n_reps <- 5000
was_it_heads <- numeric(n_reps)
for(m in 1:n_reps){
  new_flip <- sample(sides_of_coin, 1)
  was_it_heads[m] <- new_flip == "H"
}
```

To calculate the empirical probability of the coin coming up heads, we will calculate the proportion of flips in our simulation that came up heads. To do this, we can use either the `mean()` or `sum()` function; while those functions are typically used to average or sum vectors of numbers, they can also be used on Boolean vectors like `was_it_heads`. It will treat each `TRUE` as a one and each `FALSE` as a zero in the calculation.

We use each function in the code chunk below, remembering to divide the `sum()` call by the number of flips `n_rep`.

```{r coinprob}
coin_prob <- mean(was_it_heads)
coin_prob
sum(was_it_heads) / n_reps
```

As expected, we see a proportion rather close to $0.5$! Due to the randomness of the individual flips and the finite number of flips, you likely won't observe a proportion exactly equal to $0.5$. However, the closeness of the proportion to the true probability is an illustration of the frequentist phenomenon of the long-run average. If you play around with the number of flips, you'll tend to see that the longer the run, the closer the proportion to the true probability.

### Five-Card Poker Hand

Suppose you are playing some sort of card game with a standard $52$-card deck. As part of this game, you are dealt a hand of five cards, the composition of which has some bearing on your success in the game.

Provided is a vector representing the deck of cards, aptly (in my humble opinion) called `deck_of_cards`.

```{r deckofcards}
#| tidy: true
deck_of_cards <- c("AH", "2H", "3H", "4H", "5H", "6H", "7H", "8H", "9H", "10H", "JH", "QH", "KH", "AD", "2D", "3D", "4D", "5D", "6D", "7D", "8D", "9D", "10D", "JD", "QD", "KD", "AC", "2C", "3C", "4C", "5C", "6C", "7C", "8C", "9C", "10C", "JC", "QC", "KC", "AS", "2S", "3S", "4S", "5S", "6S", "7S", "8S", "9S", "10S", "JS", "QS", "KS")
```

Recall that each card in a deck has one of four "suits" (`H`earts, `D`iamonds, `C`lubs, and `Spades`) and one of four ranks (`A`ce, 2-10, `J`ack, `Q`ueen, `K`ing). In the above vector, each card is thus represented as a string (sequences of characters between quotation marks) with the (abbreviated) rank and (abbreviated) suit.

One hand of interest is called a "full house," consisting of three cards of one rank and two cards of a different rank. (For example, {4 of Hearts, 3 of Diamonds, 4 of Spades, 4 of Clubs, 3 of Spades} would be a full house, but {4 of Hearts, 3 of Diamonds, 4 of Spades, 4 of Clubs, Queen of Spades} would not be a full house.)

So, we want to calculate the probability of getting a full house!

```{r}
#| echo: FALSE
# #### Theoretical calculation
# 
# There are only so many different five-card hands (how many?), so we can turn this into a counting problem! Our goal is to calculate the following: $$\mathbb{P}(\text{full house}) = \frac{\#\text{ of possible full houses}}{\#\text{ of possible five-card hands}}.$$
# Consider the denominator first. Here, our sample space is all the different possible five-card hands one could be dealt from a $52$-card deck. **Critically, the sample space is NOT the $52$-card deck itself.**
# 
# How many different five-card hands are there? Importantly, we do **not** care about the order of particular cards in the hand. As such, the number of distinct five-card hands that can be dealt from a $52$-card deck is the number of ways to sample five items *without replacement* from a set of $52$: $$\#\text{ of possible five-card hands} = \binom{52}{5} = \frac{52!}{5! \cdot 47!} = 2,598,960.$$
# We now consider the numerator. In a full house, we observe two distinct ranks: three cards have one rank, and two have the other.
# 
# * Rank 1: There are $\binom{13}{1} = 13$ possibilities for, or ways to choose, the first rank.
#   + There are then $4$ possible cards with this rank; since the order does not matter, there are $\binom{4}{3} = 4$ ways to choose $3$ of these.
# * Rank 2: With the first rank chosen, there are now $\binom{12}{1} = 12$ ways to choose the second rank.
#   + Just like before, there are $4$ possible cards with this rank; again, since order does not matter, there are $\binom{4}{2} = 6$ ways to choose $2$ of these.
# 
# So, we have $$\#\text{ of possible full houses} = \binom{13}{1}\binom{4}{3}\binom{12}{1}\binom{4}{2} = 13 \cdot 4 \cdot 12 \cdot 6 = 3,744.$$ Therefore, we have $$\mathbb{P}(\text{full house}) = \frac{\#\text{ of possible full houses}}{\#\text{ of possible five-card hands}} =  \boxed{\frac{3744}{2598960}}\approx \boxed{0.00144.}$$
# Let's try to simulate this!
```

#### Simulation and empirical calculation

Here's a sample (hehe, see what I did there?) piece of code to simulate the dealing of a single five-card hand.

```{r sample5draw}
sample(deck_of_cards, 5)
```

As in the previous coin-flipping example, we will use a `for`-loop to simulate $5,000$ different five-card hands. (Imagine that, after dealing each hand, we re-incorporate the dealt hand into the deck and shuffle thoroughly before the next hand is dealt.) We keep track of whether each hand was a full house in other $5,000$-element vector of Booleans; below are helper functions I defined in order to check whether a particular hand was a full house.

```{r cardhelpers}
#| tidy: TRUE
split_card <- function(card){
  return(c(substr(card, 1, nchar(card) - 1), substr(card, nchar(card), nchar(card))))
}

is_full_house <- function(hand){
  ranks <- numeric(5)
  for(c in 1:5){
    ranks[c] <- split_card(hand[c])[1]
  }
  return(length(unique(ranks)) == 2)
}
```

We then have the following code for the simulation.

```{r 5drawloop}
n_reps <- 5000
was_house_full <- numeric(n_reps)
for(m in 1:n_reps){
  new_hand <- sample(deck_of_cards, 5)
  was_house_full[m] <- is_full_house(new_hand)
}
```

As before, we calculate the empirical probability using both the `mean()` function and the `sum()` function.

```{r 5drawprob}
fivecard_prob <- mean(was_house_full)
fivecard_prob
sum(was_house_full) / n_reps
```

The true probability is approximately $0.0014$ -- the long-run average simulations strike again!

## We can count on you! (You do it.)

Well, that was fun! Shame that it's over...

Wait a second. *gasp* We have two more exercises just for you to try -- hurray!

**INSTRUCTIONS:** For each of the following two exercises, your mission (should you choose to accept it) is to perform simulations from which you calculate the empirical probabilities of the specified events.

And no, the order in which you accomplish these tasks does not matter. :) Happy counting!

### The Birthday Problem

We have the following vector of strings, formatted MMMDD, wherein each entry is one of $365$ days in a non-leap year.

```{r bdays}
#| tidy: true
bdays <- c("Jan01", "Jan02", "Jan03", "Jan04", "Jan05", "Jan06", "Jan07", "Jan08", "Jan09", "Jan10", "Jan11", "Jan12", "Jan13", "Jan14", "Jan15", "Jan16", "Jan17", "Jan18", "Jan19", "Jan20", "Jan21", "Jan22", "Jan23", "Jan24", "Jan25", "Jan26", "Jan27", "Jan28", "Jan29", "Jan30", "Jan31", "Feb01", "Feb02", "Feb03", "Feb04", "Feb05", "Feb06", "Feb07", "Feb08", "Feb09", "Feb10", "Feb11", "Feb12", "Feb13", "Feb14", "Feb15", "Feb16", "Feb17", "Feb18", "Feb19", "Feb20", "Feb21", "Feb22", "Feb23", "Feb24", "Feb25", "Feb26", "Feb27", "Feb28", "Mar01", "Mar02", "Mar03", "Mar04", "Mar05", "Mar06", "Mar07", "Mar08", "Mar09", "Mar10", "Mar11", "Mar12", "Mar13", "Mar14", "Mar15", "Mar16", "Mar17", "Mar18", "Mar19", "Mar20", "Mar21", "Mar22", "Mar23", "Mar24", "Mar25", "Mar26", "Mar27", "Mar28", "Mar29", "Mar30", "Mar31", "Apr01", "Apr02", "Apr03", "Apr04", "Apr05", "Apr06", "Apr07", "Apr08", "Apr09", "Apr10", "Apr11", "Apr12", "Apr13", "Apr14", "Apr15", "Apr16", "Apr17", "Apr18", "Apr19", "Apr20", "Apr21", "Apr22", "Apr23", "Apr24", "Apr25", "Apr26", "Apr27", "Apr28", "Apr29", "Apr30", "May01", "May02", "May03", "May04", "May05", "May06", "May07", "May08", "May09", "May10", "May11", "May12", "May13", "May14", "May15", "May16", "May17", "May18", "May19", "May20", "May21", "May22", "May23", "May24", "May25", "May26", "May27", "May28", "May29", "May30", "May31", "Jun01", "Jun02", "Jun03", "Jun04", "Jun05", "Jun06", "Jun07", "Jun08", "Jun09", "Jun10", "Jun11", "Jun12", "Jun13", "Jun14", "Jun15", "Jun16", "Jun17", "Jun18", "Jun19", "Jun20", "Jun21", "Jun22", "Jun23", "Jun24", "Jun25", "Jun26", "Jun27", "Jun28", "Jun29", "Jun30", "Jul01", "Jul02", "Jul03", "Jul04", "Jul05", "Jul06", "Jul07", "Jul08", "Jul09", "Jul10", "Jul11", "Jul12", "Jul13", "Jul14", "Jul15", "Jul16", "Jul17", "Jul18", "Jul19", "Jul20", "Jul21", "Jul22", "Jul23", "Jul24", "Jul25", "Jul26", "Jul27", "Jul28", "Jul29", "Jul30", "Jul31", "Aug01", "Aug02", "Aug03", "Aug04", "Aug05", "Aug06", "Aug07", "Aug08", "Aug09", "Aug10", "Aug11", "Aug12", "Aug13", "Aug14", "Aug15", "Aug16", "Aug17", "Aug18", "Aug19", "Aug20", "Aug21", "Aug22", "Aug23", "Aug24", "Aug25", "Aug26", "Aug27", "Aug28", "Aug29", "Aug30", "Aug31", "Sep01", "Sep02", "Sep03", "Sep04", "Sep05", "Sep06", "Sep07", "Sep08", "Sep09", "Sep10", "Sep11", "Sep12", "Sep13", "Sep14", "Sep15", "Sep16", "Sep17", "Sep18", "Sep19", "Sep20", "Sep21", "Sep22", "Sep23", "Sep24", "Sep25", "Sep26", "Sep27", "Sep28", "Sep29", "Sep30", "Oct01", "Oct02", "Oct03", "Oct04", "Oct05", "Oct06", "Oct07", "Oct08", "Oct09", "Oct10", "Oct11", "Oct12", "Oct13", "Oct14", "Oct15", "Oct16", "Oct17", "Oct18", "Oct19", "Oct20", "Oct21", "Oct22", "Oct23", "Oct24", "Oct25", "Oct26", "Oct27", "Oct28", "Oct29", "Oct30", "Oct31", "Nov01", "Nov02", "Nov03", "Nov04", "Nov05", "Nov06", "Nov07", "Nov08", "Nov09", "Nov10", "Nov11", "Nov12", "Nov13", "Nov14", "Nov15", "Nov16", "Nov17", "Nov18", "Nov19", "Nov20", "Nov21", "Nov22", "Nov23", "Nov24", "Nov25", "Nov26", "Nov27", "Nov28", "Nov29", "Nov30", "Dec01", "Dec02", "Dec03", "Dec04", "Dec05", "Dec06", "Dec07", "Dec08", "Dec09", "Dec10", "Dec11", "Dec12", "Dec13", "Dec14", "Dec15", "Dec16", "Dec17", "Dec18", "Dec19", "Dec20", "Dec21", "Dec22", "Dec23", "Dec24", "Dec25", "Dec26", "Dec27", "Dec28", "Dec29", "Dec30", "Dec31")
```

We now address the Birthday Problem: Given a group of $n$ people, **what is the probability that at least** $2$ share a birthday? Consider $n = 10$, $n = 100$, and $n = 1,000$.

#### Helpful suggestions

This is yet another counting problem! For a given $n$, what does your sample space look like?

Any of the following functions may be useful in the simulation portion of this particular exercise: `unique()`, `duplicated()`, `anyDuplicated()`. To find out more, type `?<function>` in the console and the documentation will appear in the Help tab on the right-hand side of the window.

### Two-Card Poker Hand

Suppose you are dealt a two-card hand; **what is the probability that both cards are the same suit (i.e., Hearts, Diamonds, Clubs, Spades)?** (You may use the `deck_of_cards` vector defined above to simulate this counting exercise.)
